Young's Experiment

### 2020-11-08 fdtd/em.tex
Final draft

\section{Maxwell equations}\label{s2}
Maxwell equations, in SI units, are
\begin{eqnarray}
\div\vb{D} &=& \rho_f \label{s2e1} \\
\curl\vb{E} &=& -\pdv{\vb{B}}{t} \label{s2e2} \\
\div\vb{B} &=& 0 \label{s2e3} \\
\curl\vb{H} &=& \vb{J}_f + \pdv{\vb{D}}{t}, \label{s2e4}
\end{eqnarray}
where $\vb{E}$ is the electric field, $\vb{B}$ is the magnetic field,
$\vb{D}$ is the electric displacement and $\vb{H}$ is the auxilary magnetic
vector. $\rho_f$ is the density of free charges and $\vb{J}_f$ is the 
density of free current. Equation \eqref{s2e1} is Gauss' law. Equation
\eqref{s2e2} is Faraday's law. Equation \eqref{s2e3} is the mathematical
expression of the fact that there are no magnetic monopoles. Equation
\eqref{s2e4} is Ampere's law. The term $\partial\vb{D}/\partial t$ in equation
\eqref{s2e4} is called the displacement current. Adding it to Ampere's law
was Maxwell's correction to the law. We will show that this term is critical
in getting the equation of electromagnetic waves.

@@

We will illustrate FDTD for a simple one-dimensional problem. We observed
in section \ref{s2} that an electromagnetic wave is sustained because a 
changing electric field produces a magnetic field and vice versa. For sake
of simplicity, let us consider the progagation of electromagnetic waves in
vacuum. The pair of equations from which the wave equation emerges is
\begin{eqnarray}
\curl\vb{H} &=& \epsilon_0\pdv{\vb{E}}{t} \label{s5e3} \\
\curl\vb{E} &=& -\mu_0\pdv{\vb{H}}{t} \label{s5e4}
\end{eqnarray}
Let us assume that the wave propagates in the $z$-direction with the 
electric field in the $x$-direction and the $\vb{H}$ field in the
$y$-direction. Equations \eqref{s5e3} and \eqref{s5e4} simplify to
\begin{eqnarray}
\pdv{E_x}{t} &=& -\frac{1}{\epsilon_0}\pdv{H_y}{z} \label{s5e5} \\
\pdv{H_y}{t} &=& -\frac{1}{\mu_0}\pdv{E_x}{z} \label{s5e6}
\end{eqnarray}
The fields $E_x$ and $H_y$ are functions of $z$ and $t$. In this case, the
domain of the problem is a segment along the $z$-axis. We divide it in by a 
set of grid points and evaluate the fields at each one of them at successive
instants of time. The numerical solution of the problem thus consists in
evaluating $E_x(z, t)$ and $H_y(z, t)$ at a discrete set of points. We can
approximate the derivatives in equations \eqref{s5e5} and \eqref{s5e6} in
terms of central differences and write \cite{sullivan2013electromagnetic}
\begin{eqnarray}
\frac{E_x(k, n + 1/2) - E_x(k, n - 1/2)}{\Delta t} 
 &=& -\frac{H_y(k + 1/2, n) - H_y(k - 1/2, n)}{\epsilon_0\Delta x}
\label{s5e7} \\
\frac{H_y(k + 1/2, n + 1) - H_y(k + 1/2, n)}{\Delta t} 
&=&
-\frac{E_x(k + 1, n + 1/2) - E_x(k - 1, n + 1/2)}{\epsilon_0\Delta x}
\nonumber \\
 & & \label{s5e8} 
\end{eqnarray}
We rewrite these equations as
\begin{eqnarray}
E_x\left(k ,n+\frac{1}{2}\right) &=& E_x\left(k ,
n-\frac{1}{2}\right) \nonumber \\
 & & -\frac{\Delta t}{\epsilon_0\Delta x}\left(\left(
H_y\left(k + \frac{1}{2}, n\right) - H_y\left(k - \frac{1}{2}, n\right)\right)
\right) \nonumber \\
 & & \label{s5e9}
\end{eqnarray}
and
\begin{eqnarray}
H_y\left(k+\frac{1}{2},n+1\right) &=& H_y\left(k+\frac{1}{2}, n\right) 
\nonumber \\
 & & -\frac{\Delta t}{\mu_0\Delta x}\left(
E_x\left(k + 1, n + \frac{1}{2}\right) - E_x\left(k - 1, n+\frac{1}{2}\right)
\right) \nonumber \\
 & & \label{s5e10}
\end{eqnarray}
A few observations about these equations:
\begin{itemize}
\item $E_x$ is computed at space coordinates $0, 1, \ldots$ and time 
coordinates $1/2, 3/2, \ldots$.
\item $H_y$ is computed at space coordinates $1/2, 3/2, \ldots$ and time 
coordinates $0, 1, \ldots$.
\item The points at which $E_x$ is computed are interleaved between the points
at which $H_y$ is computed in both space and time grid.
\end{itemize}
When these equations are implemented in code, we compute the fields at 
\emph{all} points in the domain \emph{at a given instant of time}. We repeat
this process for the desired number of time steps.

@@

\section{Some practical matters}\label{s6}
\subsection{Absorbing boundary condition}
\subsection{Perfectly matching layer}

\section{A simple example of FDTD}\label{s7}

>>>>

\section{Maxwell equations}\label{s2}
Maxwell equations, in SI units, are
\begin{eqnarray}
\div\vb{D} &=& \rho_f \label{s2e1} \\
\curl\vb{E} &=& -\pdv{\vb{B}}{t} \label{s2e2} \\
\div\vb{B} &=& 0 \label{s2e3} \\
\curl\vb{H} &=& \vb{J}_f + \pdv{\vb{D}}{t}, \label{s2e4}
\end{eqnarray}
where $\vb{E}$ is the electric field, $\vb{B}$ is the magnetic field,
$\vb{D}$ is the electric displacement and $\vb{H}$ is the auxiliary magnetic
vector. $\rho_f$ is the density of free charges and $\vb{J}_f$ is the 
density of free current. Equation \eqref{s2e1} is Gauss' law. Equation
\eqref{s2e2} is Faraday's law. Equation \eqref{s2e3} is the mathematical
expression of the fact that there are no magnetic monopoles. Equation
\eqref{s2e4} is Ampere's law. The term $\partial\vb{D}/\partial t$ in equation
\eqref{s2e4} is called the displacement current. Adding it to Ampere's law
was Maxwell's correction to the law. We will show that this term is critical
in getting the equation of electromagnetic waves.

@@

We will illustrate FDTD for a simple one-dimensional problem. We observed
in section \ref{s2} that an electromagnetic wave is sustained because a 
changing electric field produces a magnetic field and vice versa. For sake
of simplicity, let us consider the propagation of electromagnetic waves in
vacuum. The pair of equations from which the wave equation emerges is
\begin{eqnarray}
\curl\vb{H} &=& \epsilon_0\pdv{\vb{E}}{t} \label{s5e3} \\
\curl\vb{E} &=& -\mu_0\pdv{\vb{H}}{t} \label{s5e4}
\end{eqnarray}
Let us assume that the wave propagates in the $z$-direction with the 
electric field in the $x$-direction and the $\vb{H}$ field in the
$y$-direction. Equations \eqref{s5e3} and \eqref{s5e4} simplify to
\begin{eqnarray}
\pdv{E_x}{t} &=& -\frac{1}{\epsilon_0}\pdv{H_y}{z} \label{s5e5} \\
\pdv{H_y}{t} &=& -\frac{1}{\mu_0}\pdv{E_x}{z} \label{s5e6}
\end{eqnarray}
The fields $E_x$ and $H_y$ are functions of $z$ and $t$. In this case, the
domain of the problem is a segment along the $z$-axis. We divide it in by a 
set of grid points and evaluate the fields at each one of them at successive
instants of time. The numerical solution of the problem thus consists in
evaluating $E_x(z, t)$ and $H_y(z, t)$ at a discrete set of points. We can
approximate the derivatives in equations \eqref{s5e5} and \eqref{s5e6} in
terms of central differences and write \cite{sullivan2013electromagnetic}
\begin{eqnarray}
\frac{E_x(k, n + 1/2) - E_x(k, n - 1/2)}{\Delta t} 
 &=& -\frac{H_y(k + 1/2, n) - H_y(k - 1/2, n)}{\epsilon_0\Delta x}
\label{s5e7} \\
\frac{H_y(k + 1/2, n + 1) - H_y(k + 1/2, n)}{\Delta t} 
&=&
-\frac{E_x(k + 1, n + 1/2) - E_x(k - 1, n + 1/2)}{\epsilon_0\Delta x}
\nonumber \\
 & & \label{s5e8} 
\end{eqnarray}
We rewrite these equations as
\begin{eqnarray}
E_x\left(k ,n+\frac{1}{2}\right) &=& E_x\left(k ,
n-\frac{1}{2}\right) \nonumber \\
 & & -\frac{\Delta t}{\epsilon_0\Delta x}\left(\left(
H_y\left(k + \frac{1}{2}, n\right) - H_y\left(k - \frac{1}{2}, n\right)\right)
\right) \nonumber \\
 & & \label{s5e9}
\end{eqnarray}
and
\begin{eqnarray}
H_y\left(k+\frac{1}{2},n+1\right) &=& H_y\left(k+\frac{1}{2}, n\right) 
\nonumber \\
 & & -\frac{\Delta t}{\mu_0\Delta x}\left(
E_x\left(k + 1, n + \frac{1}{2}\right) - E_x\left(k - 1, n+\frac{1}{2}\right)
\right) \nonumber \\
 & & \label{s5e10}
\end{eqnarray}
A few observations about these equations:
\begin{itemize}
\item $E_x$ is computed at space coordinates $0, 1, \ldots$ and time 
coordinates $1/2, 3/2, \ldots$.
\item $H_y$ is computed at space coordinates $1/2, 3/2, \ldots$ and time 
coordinates $0, 1, \ldots$.
\item The points at which $E_x$ is computed are interleaved between the points
at which $H_y$ is computed in both space and time grid.
\end{itemize}
When these equations are implemented in code, we compute the fields at 
\emph{all} points in the domain \emph{at a given instant of time}. We repeat
this process for the desired number of time steps.

@@

\section{Some practical matters}\label{s6}
\subsection{Absorbing boundary condition}
A partial differential equation is not completely specified without its
boundary conditions. One of the simplest boundary conditions is the 
absorbing boundary condition. It permits the waves arriving on the boundary
proceed beyond it without change. The choice of $\Delta t$ made in equation
\eqref{s5e18} dictates that a wave will take two time steps to travel from
one grid point to the next one. Therefore, we can easily implement the
absorbing boundary condition as
\begin{equation}\label{s6e1}
\tilde{E}_x(0, n) = \tilde{E}_x(1, n-2).
\end{equation}
We do not need a separate equation for the $H_y$ field because it is 
determined by the $\tilde{E}_x$ field.

\subsection{Perfectly matching layer}
It is easy to implement an absorbing boundary condition in one-dimensional
problem because there is only one direction of propagation. In two or more
dimensions it is difficult to come up with a boundary condition analogous to
equation \eqref{s6e1} without introducing spurious reflections. In order to
get over this problem, Berenger \cite{berenger1994perfectly} introduced an
artificial absorbing layer outside the boundary that strongly attenuated a
wave incident on it. However, the problem of spurious reflection does not 
vanish. Instead of the grid boundary reflecting a wave, the absorbing layer
starts reflecting. Berenger showed that it is possible to choose the material
characteristics of the absorbing layer in such a way that there are no 
reflected waves. Such an absorbing layer is called the \emph{perfectly matching
layer}. In order to derive the equations describing it, we write Maxwell
equations in two dimensions
\begin{eqnarray}
\pdv{\tilde{\vb{D}}}{t} &=& c\curl\vb{H} \label{s6e2} \\
\pdv{\vb{H}}{t} &=& -c\curl\tilde{\vb{E}} \label{s6e3} \\
\tilde{\vb{D}}(\omega) &=& \kappa^\ast(\omega)\tilde{\vb{E}}(\omega)
\label{s6e4}
\end{eqnarray}
The complex dielectric constant in equation \eqref{s6e4} is defined as
\begin{equation}\label{s6e5}
\kappa^\ast = \kappa - \frac{i\sigma}{\omega\epsilon_0}.
\end{equation}
The field $\tilde{\vb{D}}$ is related to the electric displacement as
\begin{equation}\label{s6e6}
\tilde{\vb{D}} = c\vb{D} = \frac{1}{\sqrt{\mu_0\epsilon_0}}\vb{D}.
\end{equation}
When a wave propagating in one medium with an intrinsic impedance $\eta_1$
enters another medium with intrinsic impedance $\eta_2$ then its gets
reflected by a proportion
\begin{equation}\label{s6e7}
\Gamma = \frac{\eta_2 - \eta_1}{\eta_2 + \eta_1},
\end{equation}
where
\begin{equation}\label{s6e8}
\eta_{1,2} = \sqrt{\frac{\mu_{1,2}}{\kappa_{1,2}\epsilon_0}}.
\end{equation}
If we assume that the two media are non-magnetic then $\mu_1 = \mu_2 = \mu_0$.
If we match the characteristic impedance of the absorbing layer with the
domain then equation \eqref{s6e8} guarantees that there will be no reflection.
However, we also want to ensure that the waves are strongly attenuated in
the absorbing layer. In order to find the conditions to make that happen,
we write equations \eqref{s6e2} to \eqref{s6e4} in component form. We choose 
the TM waves so that the relevant components are $\tilde{E}_z, H_x$ and $H_y$.
If we also take a Fourier transform of these equations we get
\begin{eqnarray}
i\omega\tilde{D}_x &=& c\left(\pdv{H_y}{x} - \pdv{H_x}{y}\right)\label{s6e9} \\
\tilde{D}_z(\omega) &=& \kappa^\ast(\omega)\tilde{E}_z(\omega) \label{s6e10} \\
i\omega H_x &=& -c\pdv{\tilde{E}_z}{y} \label{s6e11} \\
i\omega H_y &=&  c\pdv{\tilde{E}_z}{x} \label{s6e12}
\end{eqnarray}
We now introduce new functions $\tilde{\kappa}, \tilde{\mu}_x$ and
$\tilde{\mu}_y$ into equations \eqref{s6e9} to \eqref{s6e12} such that
\begin{eqnarray}
i\omega\tilde{\kappa}(x)\tilde{\kappa}(y)\tilde{D}_x &=& 
c\left(\pdv{H_y}{x} - \pdv{H_x}{y}\right)\label{s6e13} \\
\tilde{D}_z(\omega) &=& \kappa^\ast(\omega)\tilde{E}_z(\omega) \label{s6e14} \\
i\omega H_x\tilde{\mu}_x(x)\tilde{\mu}_x(y) &=& 
-c\pdv{\tilde{E}_z}{y} \label{s6e15} \\
i\omega H_y\tilde{\mu}_y(x)\tilde{\mu}_y(y)&=&  
c\pdv{\tilde{E}_z}{x} \label{s6e16}
\end{eqnarray}
The functions are chosen such that
\begin{eqnarray}
\Re(\tilde{\kappa}) &=& 1 \label{s6e17} \\
\Re(\tilde{\mu}_x) &=& 1 \label{s6e18} \\
\Re(\tilde{\mu}_y) &=& 1 \label{s6e19}
\end{eqnarray}
The functions $\tilde{\mu}_x$ and $\tilde{\mu}_y$ are written as
\begin{equation}\label{s6e20}
\tilde{\mu}_i = \mu_i - \frac{\tilde{\sigma}_i}{\omega\mu_0}, i = x, y
\end{equation}
and the imaginary part of $\tilde{\kappa}$ is written as
\begin{equation}\label{s6e21}
\tilde{\kappa} = -\frac{\tilde{\sigma}}{\omega\epsilon_0}.
\end{equation}
These conditions ensure the attenuation of waves in the absorbing layer
\cite{sullivan2013electromagnetic}.

\section{A simple example of FDTD}\label{s7}
In this section we will illustrate a simple example of 1-d FDTD simulation
demonstrating the propagation of a Gaussian pulse. The code is a minor 
modification of the code available in Houle and Sullivan's book 
\cite{sullivan2013electromagnetic}. We first import the necessary packages.
\begin{verbatim}
import numpy as np
import matplotlib.pyplot as plt
\end{verbatim}
Next we set up the physical parameters of the problem.
\begin{verbatim}
ke = 200
# x-component of normalized electric field.
ex = np.zeros(ke)
# y-component of magnetic field.
hy = np.zeros(ke)

kc = ke // 2  # The centre of the grid.
t0 = 40       # The mean of the Gaussian pulse.
spread = 12   # The standard deviation of the pulse.

nsteps = 100
\end{verbatim}
In the next code snippet we evaluate the fields at all grid points for a 
$100$ iterations.
\begin{verbatim}
for t in range(1, nsteps + 1):
  for k in range(1, ke):
    ex[k] +=0.5 * (hy[k - 1] - hy[k])

  ex[kc] = np.exp(-0.5 * ((t0 - t)/spread)**2)

  for k in range(ke - 1):
    hy[k] += 0.5 * (ex[k] - ex[k + 1])
\end{verbatim}
Finally, we plot the results.
\begin{verbatim}
plt.figure(figsize=(8, 8))
plt.subplot(211)
plt.plot(ex, color='k', linewidth=1)
plt.ylabel('E$_x$', fontsize='14')
plt.xticks(np.arange(0, ke + 1, step=20))
plt.xlim(0, 200)
plt.yticks(np.arange(-1, 1.2, step=1))
plt.ylim(-1.2, 1.2)
plt.text(100, 0.5, 'T = {}'.format(t), horizontalalignment='center')
plt.subplot(212)
plt.plot(hy, color='k', linewidth=1)
plt.ylabel('H$_y$', fontsize='14')
plt.xlabel('FDTD cells')
plt.xticks(np.arange(0, ke + 1, step=20))
plt.xlim(0, 200)
plt.yticks(np.arange(-1, 1.2, step=1))
plt.ylim(-1.2, 1.2)
plt.subplots_adjust(bottom=0.2, hspace=0.45)
plt.show()
\end{verbatim}
The results of the simulation are
\begin{figure}
\centering
\includegraphics[scale=0.5]{fdtd-1}
\caption{Propagation of a Gaussian pulse}\label{f2}
\end{figure}
